---
title: "INSPIRE Threshold Clustering with GO and Pfam Comparisons"
author: "Gurkan Bebek"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
params:
  database: "../database/inspire.sqlite"
  inspire_log_dir: "../reports/all_vs_all/logs"
  go_algorithm: "inspire_legacy_v1"
  inspire_threshold: 0.60
  output_dir: "../reports/clustering"
---

```{r setup, message=FALSE, warning=FALSE}
packages <- c(
  "DBI", "RSQLite", "dplyr", "tidyr", "readr", "stringr",
  "ggplot2", "igraph", "pheatmap", "patchwork"
)
missing <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Install missing packages: ", paste(missing, collapse = ", "))

library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(igraph)
library(pheatmap)
library(patchwork)

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
set.seed(20260729)
```

# Analysis plan

This report compares petals by three independent methods:

1. **INSPIRE**: normalized network-alignment scores from the C++ logs.
2. **GO**: only genes in the two petals being compared are used. Every gene is matched to its best-scoring gene in the opposite petal, followed by a symmetric best-match average.
3. **Pfam**: only genes in the two petals being compared are used. Gene-pair Pfam similarity is Jaccard similarity, followed by the same symmetric best-match average.

INSPIRE is clustered primarily by retaining direct scores greater than `r params$inspire_threshold`. Complete-linkage threshold clusters, connected components, and weighted Walktrap communities are shown. GO and Pfam are displayed with ordinary hierarchical clustering.

# Database and helper functions

```{r database}
if (!file.exists(params$database)) stop("Database not found: ", params$database)
con <- DBI::dbConnect(RSQLite::SQLite(), params$database)

if (DBI::dbGetQuery(con, "PRAGMA integrity_check")[[1]][1] != "ok") {
  stop("SQLite integrity check failed")
}
if (nrow(DBI::dbGetQuery(con, "PRAGMA foreign_key_check")) > 0) {
  stop("SQLite foreign-key check failed")
}

petals <- DBI::dbGetQuery(
  con,
  "SELECT petal, gene_count, edge_count FROM petal ORDER BY petal"
)
petal_names <- petals$petal
n_petals <- length(petal_names)

petal_genes <- DBI::dbGetQuery(
  con,
  paste(
    "SELECT pg.petal, pg.gene_symbol, gu.uniprot_accession",
    "FROM petal_gene pg",
    "JOIN gene_uniprot gu",
    "  ON gu.gene_symbol = pg.gene_symbol",
    " AND gu.is_preferred = 1",
    "ORDER BY pg.petal, pg.gene_symbol"
  )
)

if (nrow(petal_genes) != sum(petals$gene_count)) {
  stop("Not every petal gene has exactly one preferred UniProt accession")
}

petals
```

```{r helpers}
new_similarity_matrix <- function(names) {
  m <- matrix(NA_real_, length(names), length(names), dimnames = list(names, names))
  diag(m) <- 1
  m
}

validate_similarity <- function(m, label) {
  if (anyNA(m)) stop(label, " contains missing values")
  if (max(abs(m - t(m))) > 1e-10) stop(label, " is not symmetric")
  if (any(m < -1e-10 | m > 1 + 1e-10)) stop(label, " is outside [0,1]")
}

matrix_to_long <- function(m, method) {
  as.data.frame(as.table(m), stringsAsFactors = FALSE) |>
    transmute(
      petal_a = as.character(Var1),
      petal_b = as.character(Var2),
      score = as.numeric(Freq),
      method = method
    ) |>
    filter(petal_a < petal_b)
}

symmetric_best_match_average <- function(cross_scores) {
  if (anyNA(cross_scores)) stop("Cross-petal gene score matrix contains missing values")
  forward <- mean(apply(cross_scores, 1, max))
  reverse <- mean(apply(cross_scores, 2, max))
  (forward + reverse) / 2
}

adjusted_rand_index <- function(x, y) {
  tab <- table(x, y)
  choose2 <- function(z) z * (z - 1) / 2
  observed <- sum(choose2(tab))
  rows <- sum(choose2(rowSums(tab)))
  cols <- sum(choose2(colSums(tab)))
  total <- choose2(sum(tab))
  expected <- rows * cols / total
  maximum <- (rows + cols) / 2
  if (maximum == expected) return(1)
  (observed - expected) / (maximum - expected)
}
```

# 1. INSPIRE similarity matrix

```{r inspire-matrix}
log_files <- list.files(params$inspire_log_dir, pattern = "\\.log$", full.names = TRUE)
if (!length(log_files)) stop("No INSPIRE logs found")

parse_log <- function(path) {
  text <- paste(readLines(path, warn = FALSE), collapse = "\n")
  hits <- stringr::str_match_all(
    text,
    "\\$RESULT\\$\\s+(\\S+)\\s+(\\S+)\\s+norm_score:\\s+(-?[0-9.]+(?:[eE][+-]?[0-9]+)?)"
  )[[1]]
  if (!nrow(hits)) return(NULL)
  hit <- hits[nrow(hits), ]
  tibble(petal_a = hit[2], petal_b = hit[3], score = as.numeric(hit[4]))
}

inspire_pairs <- bind_rows(lapply(log_files, parse_log)) |>
  filter(petal_a != petal_b) |>
  mutate(a = pmin(petal_a, petal_b), b = pmax(petal_a, petal_b)) |>
  group_by(a, b) |>
  summarise(score = last(score), .groups = "drop")

inspire_similarity <- new_similarity_matrix(petal_names)
for (i in seq_len(nrow(inspire_pairs))) {
  a <- inspire_pairs$a[i]
  b <- inspire_pairs$b[i]
  if (a %in% petal_names && b %in% petal_names) {
    inspire_similarity[a, b] <- inspire_pairs$score[i]
    inspire_similarity[b, a] <- inspire_pairs$score[i]
  }
}
diag(inspire_similarity) <- 1
validate_similarity(inspire_similarity, "INSPIRE")

write.csv(inspire_similarity,
          file.path(params$output_dir, "similarity_inspire.csv"),
          quote = FALSE)
```

# 2. GO similarity using only the two petals' genes

GO scores are queried only for the cross-product of genes in the current petal pair. The maximum GO score is used only as a fixed normalization constant.

```{r go-matrix}
go_max <- DBI::dbGetQuery(
  con,
  "SELECT MAX(score) AS maximum FROM go_pair_score WHERE algorithm = ?",
  params = list(params$go_algorithm)
)$maximum[1]
if (!is.finite(go_max) || go_max <= 0) stop("Invalid GO normalization maximum")

get_go_cross_scores <- function(petal_a, petal_b) {
  a <- petal_genes |> filter(petal == petal_a) |>
    select(gene_a = gene_symbol, protein_a = uniprot_accession)
  b <- petal_genes |> filter(petal == petal_b) |>
    select(gene_b = gene_symbol, protein_b = uniprot_accession)

  requested <- tidyr::crossing(a, b) |>
    mutate(protein1 = pmin(protein_a, protein_b),
           protein2 = pmax(protein_a, protein_b))

  scores <- DBI::dbGetQuery(
    con,
    "SELECT protein1, protein2, score FROM go_pair_score WHERE algorithm = ?",
    params = list(params$go_algorithm)
  )

  requested <- requested |>
    left_join(scores, by = c("protein1", "protein2"))
  if (anyNA(requested$score)) {
    bad <- requested |> filter(is.na(score)) |> slice(1)
    stop("Missing GO score for ", bad$gene_a, " and ", bad$gene_b)
  }

  matrix(
    requested$score / go_max,
    nrow = nrow(a), ncol = nrow(b), byrow = FALSE,
    dimnames = list(a$gene_a, b$gene_b)
  )
}

go_similarity <- new_similarity_matrix(petal_names)
go_details <- list()
idx <- 1L
for (i in seq_along(petal_names)) {
  for (j in i:length(petal_names)) {
    a <- petal_names[i]
    b <- petal_names[j]
    if (a == b) {
      value <- 1
    } else {
      cross <- get_go_cross_scores(a, b)
      value <- symmetric_best_match_average(cross)
      go_details[[idx]] <- as.data.frame(as.table(cross)) |>
        transmute(petal_a = a, petal_b = b,
                  gene_a = as.character(Var1), gene_b = as.character(Var2),
                  normalized_go_score = as.numeric(Freq))
      idx <- idx + 1L
    }
    go_similarity[a, b] <- value
    go_similarity[b, a] <- value
  }
}
validate_similarity(go_similarity, "GO")

write.csv(go_similarity,
          file.path(params$output_dir, "similarity_go_petal_genes.csv"),
          quote = FALSE)
write_csv(bind_rows(go_details),
          file.path(params$output_dir, "go_cross_petal_gene_scores.csv"))
```

# 3. Pfam similarity using only the two petals' genes

```{r pfam-matrix}
pfam_rows <- DBI::dbGetQuery(
  con,
  "SELECT uniprot_accession, pfam_accession FROM protein_pfam ORDER BY uniprot_accession, pfam_accession"
)
pfam_by_protein <- split(pfam_rows$pfam_accession, pfam_rows$uniprot_accession)

pfam_jaccard <- function(protein_a, protein_b) {
  x <- unique(pfam_by_protein[[protein_a]])
  y <- unique(pfam_by_protein[[protein_b]])
  if (is.null(x)) x <- character()
  if (is.null(y)) y <- character()
  u <- union(x, y)
  if (!length(u)) return(0)
  length(intersect(x, y)) / length(u)
}

get_pfam_cross_scores <- function(petal_a, petal_b) {
  a <- petal_genes |> filter(petal == petal_a) |>
    select(gene_a = gene_symbol, protein_a = uniprot_accession)
  b <- petal_genes |> filter(petal == petal_b) |>
    select(gene_b = gene_symbol, protein_b = uniprot_accession)

  requested <- tidyr::crossing(a, b) |>
    rowwise() |>
    mutate(score = pfam_jaccard(protein_a, protein_b)) |>
    ungroup()

  matrix(
    requested$score,
    nrow = nrow(a), ncol = nrow(b), byrow = FALSE,
    dimnames = list(a$gene_a, b$gene_b)
  )
}

pfam_similarity <- new_similarity_matrix(petal_names)
pfam_details <- list()
idx <- 1L
for (i in seq_along(petal_names)) {
  for (j in i:length(petal_names)) {
    a <- petal_names[i]
    b <- petal_names[j]
    if (a == b) {
      value <- 1
    } else {
      cross <- get_pfam_cross_scores(a, b)
      value <- symmetric_best_match_average(cross)
      pfam_details[[idx]] <- as.data.frame(as.table(cross)) |>
        transmute(petal_a = a, petal_b = b,
                  gene_a = as.character(Var1), gene_b = as.character(Var2),
                  pfam_jaccard = as.numeric(Freq))
      idx <- idx + 1L
    }
    pfam_similarity[a, b] <- value
    pfam_similarity[b, a] <- value
  }
}
validate_similarity(pfam_similarity, "Pfam")

write.csv(pfam_similarity,
          file.path(params$output_dir, "similarity_pfam_petal_genes.csv"),
          quote = FALSE)
write_csv(bind_rows(pfam_details),
          file.path(params$output_dir, "pfam_cross_petal_gene_scores.csv"))
```

# Similarity distributions

```{r distributions, fig.width=10, fig.height=7}
all_scores <- bind_rows(
  matrix_to_long(inspire_similarity, "INSPIRE"),
  matrix_to_long(go_similarity, "GO petal genes"),
  matrix_to_long(pfam_similarity, "Pfam petal genes")
)
write_csv(all_scores, file.path(params$output_dir, "all_similarity_pairs_long.csv"))

ggplot(all_scores, aes(score)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~method, ncol = 1, scales = "free_y") +
  theme_bw() +
  labs(title = "Off-diagonal petal similarities", x = "Similarity", y = "Pairs")
```

# INSPIRE threshold clustering

## Complete-linkage clusters at the selected threshold

Complete linkage is the strict interpretation: all members of a cluster must remain close enough that the maximum within-cluster distance is no greater than `1 - threshold`.

```{r inspire-complete-linkage, fig.width=10, fig.height=6}
cutoff <- params$inspire_threshold
inspire_complete_tree <- hclust(as.dist(1 - inspire_similarity), method = "complete")
inspire_complete_clusters <- cutree(inspire_complete_tree, h = 1 - cutoff)

complete_assignments <- tibble(
  petal = names(inspire_complete_clusters),
  cluster = as.integer(inspire_complete_clusters)
) |>
  group_by(cluster) |>
  mutate(cluster_size = n()) |>
  ungroup() |>
  arrange(desc(cluster_size), cluster, petal)

write_csv(complete_assignments,
          file.path(params$output_dir, "inspire_complete_linkage_threshold_clusters.csv"))

plot(inspire_complete_tree,
     main = paste0("INSPIRE complete linkage; score cutoff = ", cutoff),
     xlab = "", sub = "", hang = -1)
abline(h = 1 - cutoff, col = "red", lty = 2, lwd = 2)

complete_assignments
```

## Thresholded heatmap

```{r inspire-threshold-heatmap, fig.width=9, fig.height=9}
threshold_matrix <- inspire_similarity
threshold_matrix[threshold_matrix <= cutoff] <- 0
diag(threshold_matrix) <- 1

order_names <- inspire_complete_tree$labels[inspire_complete_tree$order]
annotation <- data.frame(
  ThresholdCluster = factor(inspire_complete_clusters[order_names]),
  row.names = order_names
)

pheatmap(
  threshold_matrix[order_names, order_names],
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_row = annotation,
  annotation_col = annotation,
  main = paste0("INSPIRE scores above ", cutoff),
  color = colorRampPalette(c("white", "#fee8c8", "#fdbb84", "#e34a33", "#7f0000"))(100),
  border_color = NA
)
```

## Connected components and weighted Walktrap communities

```{r inspire-graph, fig.width=10, fig.height=8}
inspire_long <- matrix_to_long(inspire_similarity, "INSPIRE")
high_edges <- inspire_long |>
  filter(score > cutoff) |>
  transmute(from = petal_a, to = petal_b, weight = score)

high_graph <- graph_from_data_frame(
  high_edges,
  directed = FALSE,
  vertices = tibble(name = petal_names)
)

component_membership <- components(high_graph)$membership
component_assignments <- tibble(
  petal = names(component_membership),
  component = as.integer(component_membership)
) |>
  group_by(component) |>
  mutate(component_size = n()) |>
  ungroup() |>
  arrange(desc(component_size), component, petal)

write_csv(component_assignments,
          file.path(params$output_dir, "inspire_threshold_connected_components.csv"))

if (ecount(high_graph) > 0) {
  walktrap <- cluster_walktrap(high_graph, weights = E(high_graph)$weight)
  walktrap_membership <- membership(walktrap)
} else {
  walktrap_membership <- setNames(seq_along(petal_names), petal_names)
}

walktrap_assignments <- tibble(
  petal = names(walktrap_membership),
  cluster = as.integer(walktrap_membership)
) |>
  group_by(cluster) |>
  mutate(cluster_size = n()) |>
  ungroup() |>
  arrange(desc(cluster_size), cluster, petal)

write_csv(walktrap_assignments,
          file.path(params$output_dir, "inspire_threshold_walktrap_clusters.csv"))

set.seed(20260729)
plot(
  high_graph,
  layout = layout_with_fr(high_graph),
  vertex.color = walktrap_membership[V(high_graph)$name],
  vertex.size = 22,
  vertex.label.cex = 0.8,
  edge.width = 1 + 5 * E(high_graph)$weight,
  edge.color = adjustcolor("grey30", alpha.f = 0.55),
  main = paste0("Weighted INSPIRE graph; score > ", cutoff)
)
```

# Threshold sensitivity

```{r threshold-sensitivity, fig.width=11, fig.height=4.5}
thresholds <- seq(0.50, 0.75, by = 0.025)
threshold_sensitivity <- bind_rows(lapply(thresholds, function(t) {
  edges <- inspire_long |> filter(score > t)
  g <- graph_from_data_frame(
    edges |> transmute(from = petal_a, to = petal_b, weight = score),
    directed = FALSE,
    vertices = tibble(name = petal_names)
  )
  complete_clusters <- cutree(
    hclust(as.dist(1 - inspire_similarity), method = "complete"),
    h = 1 - t
  )
  tibble(
    threshold = t,
    retained_edges = nrow(edges),
    connected_components = components(g)$no,
    largest_component = max(components(g)$csize),
    isolated_petals = sum(degree(g) == 0),
    complete_linkage_clusters = length(unique(complete_clusters))
  )
}))

write_csv(threshold_sensitivity,
          file.path(params$output_dir, "inspire_threshold_sensitivity.csv"))

p1 <- ggplot(threshold_sensitivity, aes(threshold, retained_edges)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = cutoff, linetype = 2) +
  theme_bw() + labs(title = "Retained INSPIRE edges", y = "Edges")

p2 <- ggplot(threshold_sensitivity, aes(threshold, complete_linkage_clusters)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = cutoff, linetype = 2) +
  theme_bw() + labs(title = "Complete-linkage clusters", y = "Clusters")

p1 + p2
```

# GO and Pfam hierarchical clustering

```{r go-pfam-hclust, fig.width=12, fig.height=9}
go_tree <- hclust(as.dist(1 - go_similarity), method = "average")
pfam_tree <- hclust(as.dist(1 - pfam_similarity), method = "average")

old_par <- par(no.readonly = TRUE)
par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))
plot(go_tree, main = "GO petal-gene hierarchical clustering", xlab = "", sub = "")
plot(pfam_tree, main = "Pfam petal-gene hierarchical clustering", xlab = "", sub = "")
par(old_par)
```

```{r go-heatmap, fig.width=9, fig.height=9}
pheatmap(
  go_similarity,
  clustering_distance_rows = as.dist(1 - go_similarity),
  clustering_distance_cols = as.dist(1 - go_similarity),
  clustering_method = "average",
  main = "GO petal-gene similarity",
  border_color = NA
)
```

```{r pfam-heatmap, fig.width=9, fig.height=9}
pheatmap(
  pfam_similarity,
  clustering_distance_rows = as.dist(1 - pfam_similarity),
  clustering_distance_cols = as.dist(1 - pfam_similarity),
  clustering_method = "average",
  main = "Pfam petal-gene similarity",
  border_color = NA
)
```

# Final INSPIRE threshold-cluster comparison

```{r final-cluster-table}
final_cluster_table <- tibble(
  petal = petal_names,
  complete_linkage_cluster = as.integer(inspire_complete_clusters[petal_names]),
  connected_component = as.integer(component_membership[petal_names]),
  walktrap_cluster = as.integer(walktrap_membership[petal_names])
)

write_csv(final_cluster_table,
          file.path(params$output_dir, "inspire_threshold_cluster_comparison.csv"))

final_cluster_table
```

# Matrix correlations

```{r correlations}
all_wide <- all_scores |>
  select(petal_a, petal_b, method, score) |>
  pivot_wider(names_from = method, values_from = score)

pearson <- cor(all_wide |> select(where(is.numeric)), method = "pearson")
spearman <- cor(all_wide |> select(where(is.numeric)), method = "spearman")

write.csv(pearson, file.path(params$output_dir, "matrix_correlations_pearson.csv"))
write.csv(spearman, file.path(params$output_dir, "matrix_correlations_spearman.csv"))

pearson
spearman
```

# Database metadata

```{r metadata}
metadata <- DBI::dbGetQuery(con, "SELECT key, value FROM metadata ORDER BY key")
write_csv(metadata, file.path(params$output_dir, "database_metadata.csv"))
metadata
```

# Disconnect

```{r disconnect}
DBI::dbDisconnect(con)
```

# Reproducibility

```{r reproducibility}
sessionInfo()
```
