---
title: "INSPIRE Human Network Clustering"
author: "Gurkan Bebek"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
params:
  database: "../database/inspire.sqlite"
  inspire_log_dir: "../reports/all_vs_all/logs"
  go_algorithm: "inspire_legacy_v1"
  expected_go_normalization: 14.1026
  threshold_min: 0.45
  threshold_max: 0.70
  threshold_step: 0.025
  selected_threshold: 0.60
  output_dir: "../reports/clustering_current_human"
---

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "DBI", "RSQLite", "dplyr", "tidyr", "readr", "stringr",
  "ggplot2", "igraph", "pheatmap", "patchwork"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop("Install missing packages: ", paste(missing_packages, collapse = ", "))
}

library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(igraph)
library(pheatmap)
library(patchwork)

set.seed(20260729)
dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

# Purpose

This report analyzes the current human INSPIRE results after validating the executable with the GO normalization associated with the current GO cache.

The primary analysis uses the complete INSPIRE similarity matrix and hierarchical clustering. Two linkage methods are shown:

1. **Complete linkage**, as described in the manuscript Methods.
2. **Ward.D2**, because the historical figure command displayed Ward clustering.

Thresholded networks from 0.45 through 0.70 are secondary sensitivity analyses. They are not used to choose or manufacture a preferred primary clustering.

# Database validation

```{r database}
if (!file.exists(params$database)) {
  stop("Database not found: ", params$database)
}

con <- DBI::dbConnect(RSQLite::SQLite(), params$database)

integrity <- DBI::dbGetQuery(con, "PRAGMA integrity_check")[[1]][1]
if (!identical(integrity, "ok")) {
  stop("SQLite integrity check failed: ", integrity)
}

foreign_key_errors <- DBI::dbGetQuery(con, "PRAGMA foreign_key_check")
if (nrow(foreign_key_errors) > 0) {
  stop("SQLite foreign-key check returned errors")
}

petals <- DBI::dbGetQuery(
  con,
  "SELECT petal, gene_count, edge_count FROM petal ORDER BY petal"
)

petal_names <- petals$petal
n_petals <- length(petal_names)

if (n_petals < 3) stop("At least three petals are required")

petals
```

# GO-cache normalization audit

```{r normalization-audit}
go_audit <- DBI::dbGetQuery(
  con,
  paste(
    "SELECT algorithm,",
    "       COUNT(*) AS cached_pairs,",
    "       MIN(score) AS minimum_score,",
    "       MAX(score) AS maximum_score,",
    "       AVG(score) AS mean_score",
    "FROM go_pair_score",
    "WHERE algorithm = ?",
    "GROUP BY algorithm"
  ),
  params = list(params$go_algorithm)
)

if (nrow(go_audit) != 1) {
  stop("No GO cache was found for algorithm: ", params$go_algorithm)
}

normalization_difference <- abs(
  go_audit$maximum_score[1] - params$expected_go_normalization
)

if (normalization_difference > 1e-4) {
  stop(
    "GO-cache maximum is ", go_audit$maximum_score[1],
    " but expected_go_normalization is ", params$expected_go_normalization,
    ". Update the parameter or regenerate the correct logs."
  )
}

go_audit
```

# Read completed INSPIRE logs

```{r inspire-logs}
if (!dir.exists(params$inspire_log_dir)) {
  stop("INSPIRE log directory not found: ", params$inspire_log_dir)
}

log_files <- list.files(
  params$inspire_log_dir,
  pattern = "\\.log$|\\.txt$",
  full.names = TRUE
)

if (length(log_files) == 0) {
  stop("No INSPIRE .log or .txt files found in ", params$inspire_log_dir)
}

parse_inspire_log <- function(path) {
  lines <- readLines(path, warn = FALSE)
  result_lines <- grep("\\$RESULT\\$", lines, value = TRUE)
  normalization_lines <- grep("GO normalization in use:", lines, value = TRUE)

  normalization <- NA_real_
  if (length(normalization_lines) > 0) {
    normalization <- suppressWarnings(as.numeric(
      stringr::str_extract(tail(normalization_lines, 1), "[0-9]+(?:\\.[0-9]+)?")
    ))
  }

  if (length(result_lines) == 0) {
    return(tibble(
      log_file = path,
      petal_a = NA_character_,
      petal_b = NA_character_,
      score = NA_real_,
      normalization = normalization,
      status = "incomplete"
    ))
  }

  last_result <- tail(result_lines, 1)
  fields <- stringr::str_split(stringr::str_squish(last_result), "\\s+")[[1]]

  score_index <- which(fields == "norm_score:")
  if (length(score_index) != 1 || score_index + 1 > length(fields)) {
    return(tibble(
      log_file = path,
      petal_a = NA_character_,
      petal_b = NA_character_,
      score = NA_real_,
      normalization = normalization,
      status = "unparseable"
    ))
  }

  tibble(
    log_file = path,
    petal_a = fields[2],
    petal_b = fields[3],
    score = suppressWarnings(as.numeric(fields[score_index + 1])),
    normalization = normalization,
    status = "complete"
  )
}

log_results <- bind_rows(lapply(log_files, parse_inspire_log))

log_audit <- log_results |>
  count(status, name = "files")

log_audit
```

```{r log-normalization-check}
reported_normalizations <- log_results |>
  filter(status == "complete", is.finite(normalization)) |>
  distinct(normalization) |>
  pull(normalization)

if (length(reported_normalizations) > 0 &&
    any(abs(reported_normalizations - params$expected_go_normalization) > 1e-4)) {
  stop(
    "At least one log used a different GO normalization: ",
    paste(reported_normalizations, collapse = ", ")
  )
}

incomplete_logs <- log_results |>
  filter(status != "complete" | !is.finite(score))

if (nrow(incomplete_logs) > 0) {
  write_csv(
    incomplete_logs,
    file.path(params$output_dir, "incomplete_or_unparseable_logs.csv")
  )
}

reported_normalizations
```

# Build and validate the similarity matrix

```{r similarity-matrix}
pair_results <- log_results |>
  filter(
    status == "complete",
    is.finite(score),
    petal_a %in% petal_names,
    petal_b %in% petal_names,
    petal_a != petal_b
  ) |>
  mutate(
    first = pmin(petal_a, petal_b),
    second = pmax(petal_a, petal_b)
  ) |>
  group_by(first, second) |>
  summarise(
    score = last(score),
    files = n(),
    .groups = "drop"
  )

expected_pairs <- choose(n_petals, 2)
observed_pairs <- nrow(pair_results)

pair_coverage <- tibble(
  petals = n_petals,
  expected_off_diagonal_pairs = expected_pairs,
  observed_off_diagonal_pairs = observed_pairs,
  coverage = observed_pairs / expected_pairs
)

pair_coverage
```

```{r missing-pairs}
expected_pair_table <- combn(petal_names, 2, simplify = FALSE) |>
  lapply(function(x) tibble(first = min(x), second = max(x))) |>
  bind_rows()

missing_pairs <- expected_pair_table |>
  anti_join(pair_results, by = c("first", "second"))

if (nrow(missing_pairs) > 0) {
  write_csv(
    missing_pairs,
    file.path(params$output_dir, "missing_inspire_pairs.csv")
  )

  stop(
    "INSPIRE matrix is incomplete: ", nrow(missing_pairs),
    " of ", expected_pairs, " off-diagonal pairs are missing. See missing_inspire_pairs.csv."
  )
}

inspire_similarity <- matrix(
  NA_real_,
  nrow = n_petals,
  ncol = n_petals,
  dimnames = list(petal_names, petal_names)
)

diag(inspire_similarity) <- 1

for (i in seq_len(nrow(pair_results))) {
  a <- pair_results$first[i]
  b <- pair_results$second[i]
  inspire_similarity[a, b] <- pair_results$score[i]
  inspire_similarity[b, a] <- pair_results$score[i]
}

if (anyNA(inspire_similarity)) stop("INSPIRE matrix contains missing values")
if (max(abs(inspire_similarity - t(inspire_similarity))) > 1e-10) {
  stop("INSPIRE matrix is not symmetric")
}
if (any(inspire_similarity < 0 | inspire_similarity > 1)) {
  stop("INSPIRE matrix contains values outside [0,1]")
}

write.csv(
  inspire_similarity,
  file.path(params$output_dir, "similarity_inspire_current_human.csv"),
  quote = FALSE
)
```

# Positive-control scores

```{r positive-controls}
get_pair_score <- function(a, b, similarity) {
  if (!a %in% rownames(similarity) || !b %in% colnames(similarity)) return(NA_real_)
  similarity[a, b]
}

positive_controls <- tibble(
  pair = c("SMAD2-SMAD3", "SMAD2-SMAD4", "SMAD3-SMAD4", "NF1-KRAS"),
  petal_a = c("SMAD2", "SMAD2", "SMAD3", "NF1"),
  petal_b = c("SMAD3", "SMAD4", "SMAD4", "KRAS")
) |>
  rowwise() |>
  mutate(score = get_pair_score(petal_a, petal_b, inspire_similarity)) |>
  ungroup()

all_off_diagonal <- inspire_similarity[upper.tri(inspire_similarity)]

positive_controls <- positive_controls |>
  mutate(
    percentile_among_all_pairs = vapply(
      score,
      function(x) if (is.na(x)) NA_real_ else mean(all_off_diagonal <= x),
      numeric(1)
    )
  )

write_csv(
  positive_controls,
  file.path(params$output_dir, "positive_control_scores.csv")
)

positive_controls
```

# Score distribution

```{r score-distribution, fig.width=8, fig.height=5}
score_long <- pair_results |>
  transmute(petal_a = first, petal_b = second, score = score)

score_summary <- score_long |>
  summarise(
    pairs = n(),
    minimum = min(score),
    first_quartile = quantile(score, 0.25),
    median = median(score),
    mean = mean(score),
    third_quartile = quantile(score, 0.75),
    maximum = max(score),
    above_0_45 = sum(score > 0.45),
    above_0_50 = sum(score > 0.50),
    above_0_60 = sum(score > 0.60),
    above_0_70 = sum(score > 0.70)
  )

write_csv(
  score_summary,
  file.path(params$output_dir, "inspire_score_summary.csv")
)

ggplot(score_long, aes(score)) +
  geom_histogram(binwidth = 0.025, boundary = 0, color = "white") +
  geom_vline(
    xintercept = c(0.45, 0.50, 0.60, 0.70),
    linetype = c(3, 3, 2, 3)
  ) +
  theme_bw() +
  labs(
    title = "Current human INSPIRE similarity distribution",
    x = "Normalized INSPIRE score",
    y = "Petal pairs"
  )

score_summary
```

# Primary hierarchical clustering

```{r hierarchical-clustering}
inspire_distance <- as.dist(1 - inspire_similarity)

hc_complete <- hclust(
  inspire_distance,
  method = "complete"
)

hc_ward <- hclust(
  inspire_distance,
  method = "ward.D2"
)
```

```{r dendrograms, fig.width=13, fig.height=10}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(2, 1), mar = c(4, 4, 4, 1))

plot(
  hc_complete,
  main = "INSPIRE hierarchical clustering: complete linkage",
  xlab = "",
  sub = "",
  hang = -1
)

plot(
  hc_ward,
  main = "INSPIRE hierarchical clustering: Ward.D2",
  xlab = "",
  sub = "",
  hang = -1
)

par(old_par)
```

```{r complete-heatmap, fig.width=10, fig.height=10}
pheatmap(
  inspire_similarity,
  clustering_distance_rows = inspire_distance,
  clustering_distance_cols = inspire_distance,
  clustering_method = "complete",
  main = "Current human INSPIRE similarity: complete linkage",
  border_color = NA
)
```

```{r ward-heatmap, fig.width=10, fig.height=10}
pheatmap(
  inspire_similarity,
  clustering_distance_rows = inspire_distance,
  clustering_distance_cols = inspire_distance,
  clustering_method = "ward.D2",
  main = "Current human INSPIRE similarity: Ward.D2",
  border_color = NA
)
```

# Inspect cluster solutions without selecting a single k

```{r cluster-solutions}
k_values <- 2:min(10, n_petals - 1)

cluster_solutions <- bind_rows(lapply(k_values, function(k) {
  tibble(
    petal = petal_names,
    k = k,
    complete_cluster = unname(cutree(hc_complete, k = k)[petal_names]),
    ward_cluster = unname(cutree(hc_ward, k = k)[petal_names])
  )
}))

write_csv(
  cluster_solutions,
  file.path(params$output_dir, "hierarchical_cluster_solutions_k2_to_k10.csv")
)

cluster_solutions
```

# Secondary threshold sensitivity analysis

```{r threshold-grid}
thresholds <- seq(
  params$threshold_min,
  params$threshold_max,
  by = params$threshold_step
)

threshold_rows <- vector("list", length(thresholds))
component_rows <- vector("list", length(thresholds))
walktrap_rows <- vector("list", length(thresholds))

for (index in seq_along(thresholds)) {
  threshold <- thresholds[index]

  retained_edges <- score_long |>
    filter(score > threshold) |>
    transmute(from = petal_a, to = petal_b, weight = score)

  graph <- graph_from_data_frame(
    retained_edges,
    directed = FALSE,
    vertices = tibble(name = petal_names)
  )

  components_result <- components(graph)

  if (ecount(graph) > 0) {
    walktrap_result <- cluster_walktrap(graph, weights = E(graph)$weight)
    walktrap_membership <- membership(walktrap_result)
  } else {
    walktrap_membership <- setNames(seq_along(petal_names), petal_names)
  }

  threshold_rows[[index]] <- tibble(
    threshold = threshold,
    retained_edges = nrow(retained_edges),
    retained_fraction = nrow(retained_edges) / nrow(score_long),
    isolated_petals = sum(degree(graph) == 0),
    connected_components = components_result$no,
    largest_component = max(components_result$csize),
    walktrap_communities = length(unique(walktrap_membership))
  )

  component_rows[[index]] <- tibble(
    threshold = threshold,
    petal = petal_names,
    component = unname(components_result$membership[petal_names])
  )

  walktrap_rows[[index]] <- tibble(
    threshold = threshold,
    petal = petal_names,
    community = unname(walktrap_membership[petal_names])
  )
}

threshold_summary <- bind_rows(threshold_rows)
component_assignments <- bind_rows(component_rows)
walktrap_assignments <- bind_rows(walktrap_rows)

write_csv(
  threshold_summary,
  file.path(params$output_dir, "threshold_grid_summary.csv")
)
write_csv(
  component_assignments,
  file.path(params$output_dir, "threshold_component_assignments.csv")
)
write_csv(
  walktrap_assignments,
  file.path(params$output_dir, "threshold_walktrap_assignments.csv")
)

threshold_summary
```

```{r threshold-grid-plots, fig.width=11, fig.height=8}
p_edges <- ggplot(threshold_summary, aes(threshold, retained_edges)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = params$selected_threshold, linetype = 2) +
  theme_bw() +
  labs(title = "Retained INSPIRE edges", y = "Edges")

p_isolates <- ggplot(threshold_summary, aes(threshold, isolated_petals)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = params$selected_threshold, linetype = 2) +
  theme_bw() +
  labs(title = "Isolated petals", y = "Isolates")

p_components <- ggplot(threshold_summary, aes(threshold, connected_components)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = params$selected_threshold, linetype = 2) +
  theme_bw() +
  labs(title = "Connected components", y = "Components")

p_edges + p_isolates + p_components + plot_layout(ncol = 1)
```

# Threshold network gallery

All panels use the same node coordinates. This allows direct visual comparison as edges disappear with increasing thresholds.

```{r threshold-network-gallery, results='asis'}
lowest_threshold <- min(thresholds)
lowest_edges <- score_long |>
  filter(score > lowest_threshold) |>
  transmute(from = petal_a, to = petal_b, weight = score)

layout_graph <- graph_from_data_frame(
  lowest_edges,
  directed = FALSE,
  vertices = tibble(name = petal_names)
)

set.seed(20260729)
fixed_layout <- layout_with_fr(
  layout_graph,
  weights = E(layout_graph)$weight,
  niter = 2000
)
rownames(fixed_layout) <- V(layout_graph)$name
fixed_layout <- fixed_layout[petal_names, , drop = FALSE]

network_dir <- file.path(params$output_dir, "threshold_networks")
dir.create(network_dir, recursive = TRUE, showWarnings = FALSE)

pdf_path <- file.path(params$output_dir, "threshold_networks_all.pdf")
pdf(pdf_path, width = 10, height = 8, onefile = TRUE)

for (index in seq_along(thresholds)) {
  threshold <- thresholds[index]

  retained_edges <- score_long |>
    filter(score > threshold) |>
    transmute(from = petal_a, to = petal_b, weight = score)

  graph <- graph_from_data_frame(
    retained_edges,
    directed = FALSE,
    vertices = tibble(name = petal_names)
  )

  if (ecount(graph) > 0) {
    communities <- membership(cluster_walktrap(graph, weights = E(graph)$weight))
  } else {
    communities <- setNames(seq_along(petal_names), petal_names)
  }

  palette <- hcl.colors(max(communities), palette = "Dark 3")
  node_colors <- palette[communities[V(graph)$name]]
  node_degree <- degree(graph)
  node_shapes <- ifelse(node_degree == 0, "square", "circle")
  node_frames <- ifelse(node_degree == 0, "black", "grey30")

  edge_widths <- if (ecount(graph) > 0) {
    1 + 5 * (E(graph)$weight - threshold) / max(1e-12, 1 - threshold)
  } else {
    numeric(0)
  }

  title_text <- paste0(
    "INSPIRE > ", sprintf("%.3f", threshold),
    " | edges: ", ecount(graph),
    " | components: ", components(graph)$no,
    " | isolates: ", sum(node_degree == 0)
  )

  draw_graph <- function() {
    plot(
      graph,
      layout = fixed_layout[V(graph)$name, , drop = FALSE],
      vertex.color = node_colors,
      vertex.frame.color = node_frames,
      vertex.shape = node_shapes,
      vertex.size = 24,
      vertex.label.cex = 0.8,
      edge.width = edge_widths,
      edge.color = adjustcolor("grey35", alpha.f = 0.55),
      main = title_text,
      margin = 0.12
    )
  }

  draw_graph()

  png_path <- file.path(
    network_dir,
    paste0("INSPIRE_threshold_", sprintf("%.3f", threshold), ".png")
  )

  png(png_path, width = 1800, height = 1400, res = 180)
  draw_graph()
  dev.off()

  cat("\n\n## Threshold ", sprintf("%.3f", threshold), "\n\n", sep = "")
  print(knitr::include_graphics(png_path))
  cat("\n\n")
}

dev.off()
```

# Selected-threshold assignments

```{r selected-threshold}
selected_index <- which.min(abs(thresholds - params$selected_threshold))
selected_threshold <- thresholds[selected_index]

selected_components <- component_assignments |>
  filter(abs(threshold - selected_threshold) < 1e-10) |>
  select(petal, component)

selected_walktrap <- walktrap_assignments |>
  filter(abs(threshold - selected_threshold) < 1e-10) |>
  select(petal, community)

selected_assignments <- selected_components |>
  left_join(selected_walktrap, by = "petal") |>
  arrange(component, community, petal)

write_csv(
  selected_assignments,
  file.path(params$output_dir, "selected_threshold_assignments.csv")
)

selected_assignments
```

# Analysis record

```{r analysis-record}
analysis_record <- tibble(
  item = c(
    "GO algorithm",
    "GO normalization",
    "Petals",
    "Expected off-diagonal pairs",
    "Observed off-diagonal pairs",
    "Primary clustering 1",
    "Primary clustering 2",
    "Threshold analysis role"
  ),
  value = c(
    params$go_algorithm,
    as.character(params$expected_go_normalization),
    as.character(n_petals),
    as.character(expected_pairs),
    as.character(observed_pairs),
    "complete linkage on 1 - INSPIRE similarity",
    "Ward.D2 on 1 - INSPIRE similarity",
    "secondary sensitivity analysis"
  )
)

write_csv(
  analysis_record,
  file.path(params$output_dir, "analysis_record.csv")
)

analysis_record
```

# Database metadata

```{r metadata}
metadata <- DBI::dbGetQuery(
  con,
  "SELECT key, value FROM metadata ORDER BY key"
)

write_csv(
  metadata,
  file.path(params$output_dir, "database_metadata.csv")
)

metadata
```

# Disconnect

```{r disconnect}
DBI::dbDisconnect(con)
```

# Reproducibility

```{r reproducibility}
sessionInfo()
```
